# Data types and data structures

In this chapter, we will cover the following topics -

* Data type
  + Numbers
  + Characters
  + Logical
* Data structure
  + Vector
  + Matrix
  + Array
  + Data frame
  + List
* Import and export data
  + .txt
  + .csv
  + .xlsx
* Manipulating data


## Useful data types in *R*

There are different data types used in *R*. We will touch upon a few but heavily used data types in this class. These are namely - 


### Numbers: 
Numbers can be an Integers, or can be floating numbers (decimal numbers). They are recognized as simple `Numeric` data type in *R*. 

###  Character: 
This type of data can be equivalent to categorical variable. In other programming language, this is equivalent to *String* or *Text* data type. Here in *R*, these can be nominal (just names) or ordinal (names with hidden order) variables. 

### Logical: 
such as True or False. These can be the result of a boolean operation. 

Let’s see an example - 

```{r, echo=F, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)
kable(
  data.frame(
    "participant ID"=c("BMI1_001", "BMI1_002", "BMI1_003"),
    Date=c("01/11/2021", "02/11/2021", "03/11/2021"), 
    "weight in kg"=c(55.5,45,75), "height in cm"=c(172,165,150),
    smoker=c("No", "Yes", "No"),
    "Overall health-condition"=c("Intermediate", "Good", "Poor")
    ), 
  format = "html") %>%
  kable_styling(
    bootstrap_options = "striped", 
    full_width = F, 
    position = "left"
    )
```
Here in the table above, *participant.ID*, *smoker*, *Overall.health.condition* columns represent character data type. Of them, *participant.ID* and *smoker* columns contain nominal data, while the *Overall.health.condition* represents the ordinal data type.

The columns namely *weight.in.kg* and	*height.in.cm* contain numeric data type. 

A very very important information - 
Categorical data (nominal or ordinal) in *R* are also called *Factors*. Sometimes, you need to detach this characteristic (i.e. being factors) from your data and convert it to merely character type for the sake of data handling. 

```{r, results= F}
Overall.health.condition <- factor(c("Intermediate", "Good", "Poor", "Good", "Poor" ))
Overall.health.condition
#~~~~~~~~~~~~~~~~~~~~
Overall.health.condition <- factor(c("Intermediate", "Good", "Poor", "Good", "Poor"), 
                                   ordered = F)
Overall.health.condition
#~~~~~~~~~~~~~~~~~~~~
Overall.health.condition <- factor(c("Intermediate", "Good", "Poor", "Good", "Poor"), 
                                   ordered = T)
Overall.health.condition
#~~~~~~~~~~~~~~~~~~~~
Overall.health.condition <- factor(c("Intermediate", "Good", "Poor", "Good", "Poor"), 
                                   ordered = T, 
                                   levels = c("Poor", "Intermediate", "Good"))
Overall.health.condition
```

There are also *complex* data types, *date* data types etc, which we will not cover in this course.


## Data structure

There are few useful pre-defined data structures heavily used in *R* and there are user defined data structures as well (there are no limits in this case). We will discuss a few of the former class -


### Vector: 
Let's first think of a variable and value pair, like you have a variable `x` and it has a value of `2`. Like what we write in *R*: 
`x <- 2`, which means, we assign the value `2` to `x`. This is equivalent to a *scalar*. 
You can put more than one value in a variable, and then it will be equivalent to a *vector*: 
```
y <- c(1,2,3,4,5,6)
``` 
Here we have used the *combine function* `c()` to create a vector. Vector can hold different data-types individually, that means, one vector can hold only one type of data (numeric or character or logical etc). Just remember, if you want to hold multiple elements (be it numeric values or characters), enclose them with the *combine function* (as above).

Each of the component of the vector is called its *Element*. You can call or refer the elements individually or as a group using their *positional index*. The positional index starts from `1`, which you can refer to as *one-based indexing*. For example, if you want to call or retrieve the first element of vector `y`, you can type in `y[1]`. Similarly, you can refer to the 4th element by typing in `y[4]`. If you want to refer to from 2nd to the 4th elements of `y`, you type in `y[c(2,3,4)]` or you can use `y[2:4]`. 

A useful tip - for sequential numbers, you can use a short-cut using colon sign in between the minimum and maximum values (inclusive) of the range. Like, for 1 to 100, you type in `1:100`.

#### Short exercise on *Vector*

Create a vector called `my_vec` and assign it with letters `A b c D e`. Then call the vector on the Console.


### Matrix: 
A matrix is a two-dimensional (row and column) array of a single data type. Matrices are created using `matrix()` function. Think of you having 20 numbers, from 1 to 20 and you want to create a matrix with 4 rows and 5 columns. So you type in:
```{r}
myMatrix <- matrix(data=1:20,nrow = 4,ncol = 5)
myMatrix
```
If you want to subset the matrix you can do it easily. Always remember, matrix is a `m` row by `n` column `(m x n)` structure which you define in *R* as `[m,n]` or `[row, column]`. If you want to choose the value in the first row and the first column (which also indicates numeric `1` in myMatrix), you type in `myMatrix[1,1]`. If you want to select first 2 rows and 2nd to 4th columns, you type in `myMatrix[c(1,2), c(2,3,4)]` or `myMatrix[1:2, 2:4]`.

#### Short exercise on *Matrix*

1. Create a vector called `english_alphabets` with all 26 english alphabets.[Hint: `letter()` function returns all 26 english alphabets in lower-case] 

2. Create a matrix with 13 rows and 2 columns that accommodates all those alphabets.

### Array: 
You can accommodate your data in more than 2 dimensions using an array. We will not cover it in this course.


### Data frame: 
For you, data frame is the main data type that will have a vast utility in this course. You will import data in this format and manipulate data and export data in this format. Therefore, it's imperative that you understand it very carefully. A data frame is a two-dimensional array. It is more general than a matrix and can contain different data types in it at the same time. However, each column contains a single data type (of course). The function to create a data frame is `data.frame()` and an easy way to create a data frame is 
```
myDataFrame <- data.frame(column1, column2, column3))
```
Let's see the same data frame again and observe carefully how it is done. 

```{r}
participant.ID=c("BMI1_001", "BMI1_002", "BMI1_003")
Date=c("01/11/2021", "01/11/2021", "01/11/2021")
weight.in.kg=c(55.5,45,75)
height.in.cm=c(172,165,150)
smoker=c("No", "Yes", "No")
Overall.health.condition=c("Intermediate", "Good", "Poor")

myDataFrame <- data.frame(participant.ID, Date, weight.in.kg, height.in.cm, smoker, Overall.health.condition)

myDataFrame
```

Remember, each of the vector defining each column in the data frame has to be of same length. Now, if you want to subset the data frame as you did from matrix, you can use the same technique. For example, if you want to know the weight and height of first two patients, you type in `myDataFrame[c(1,2),c(3,4)]`. Interestingly in the case of data frame, you can use the column names to do that, like -
```
myDataFrame[c(1,2),c("weight.in.kg", "height.in.cm")]
``` 
Further more, there is another level of flexibility for selecting a column information by using `dataframe$columnName` notation as in our case `myDataFrame$weight.in.kg`.

Now, you can inquire what are the column names by utilizing `colnames()` function and for the row names - `rownames()`. Here, in this case, we see there is no row name, rather they are indexed.

```{r}
rownames(myDataFrame)
```


we don't have row names yet, but we can assign them with `row1`, `row2` and `row2` by typing in 
```{r}
rownames(myDataFrame) <-  c("row1", "row2", "row3")
```

For simplicity, we will change the column names as well to simply `column1`, `column2` etc -

```{r}
colnames(myDataFrame) <- c("column1","column2","column3","column4","column5","column6")
```

And, let's see the current look of our data-frame -

```{r}
myDataFrame
```


Now, if we want to subset using row and column names, we can type in -
```{r}
subDataFrame <- myDataFrame[c("row1","row2"),c("column2", "column3")]
subDataFrame
```

Important nomenclature alert: In the realm of statistics, we refer a row as an *observation* and a column as a *variable.* In the realm of data analysts, they are *records* and *fields*, respectively. And, in the realm of Machine Learning, they are *examples* and *attributes*, respectively.

#### Short exercise on *Data Frame*

1. Re-create the same data-frame with a different name (of your choice) with the following vectors -

```{r}
participant.ID=c("BMI1_001", "BMI1_002", "BMI1_003")
Date=c("01/11/2021", "01/11/2021", "01/11/2021")
weight.in.kg=c(55.5,45,75)
height.in.cm=c(172,165,150)
smoker=c("No", "Yes", "No")
Overall.health.condition=c("Intermediate", "Good", "Poor")



```

2. Subset the columns `weight.in.kg` and `smoker` with -
  a. indexing and
  b. column names

### List: 
In *R*, List is the (arguably) the most complex data structure which can hold different data types and data structures. Even, a list can hold another list. We will not cover this data structure in this course. 


## Import and export data

For this course, we will limit ourselves to import data from our personal computers. Before that, go to the following link https://doi.org/10.5281/zenodo.6452121, download an excel file called `excell_iris_data.xlsx` and save to your working directory. Do the same for the csv file as well.

The basic mode of reading a tabular dataset is the `read.table()` function. There are few others, like read.csv(), read.csv2(), read.ftable(), read.delim() etc. There are some basic parameter that you should be careful on -

1. file = /path/to/the/file
2. header = the default is set to `TRUE`, however, if the data doesn't have a column name, set it to false.
3. sep = for `read.table()`, it is set to white space. For `read.csv()`, it is a comma. For `read.csv2()`, it's set to a semi-colon. For `read.delim()` or `read.delim2()`, it's is a tab `"\t"`.
4. row.names = if there's a column for row name (usually the first column in the raw data), specify with the column number and later get rid of that column using say, data[,column_number] <- NULL


Now we will read the csv file -
```{r}
df_csv <- read.csv(file = "csv_iris_data.csv", row.names = 1)
head(df_csv)
```
Now that we have read our data-frame and checked few rows from the top of it, we want to export it to our computer -

```{r}
write.csv(x = df_csv, file = "My_output_df.csv")
```


In the case of reading an excel file, either with *.xls* or .*xlsx* extension, we need to load a package called `readxl`.
```{r, message=FALSE, warning=FALSE}
library(readxl)
df_excel <- read_excel('excell_iris_data.xlsx',sheet='Sheet1')
head(df_excel)
```
Unfortunately, to write an excel file, we need another package -
```{r}
library(xlsx)
write.xlsx(x=df_excel, "My_output_df.xlsx")
```


## Manipulating data

First thing, I would do to check after loading a data-frame to the *R* environment is to check its dimension, look at the top and bottom of the dataset, look at the structure and summary of the data by following set of commands -

```{r}
# check the dimension of the data
dim(df_csv)
# check the head or tail
head(df_csv)
tail(df_csv)
# apply str() and summary() function 
str(df_csv)
summary(df_csv)
```

We have covered how to change the row or column names in the section `Data frame`

Now we will see how to add new column or row to the data-frame. As we have seen that there are 150 rows in `df_csv` data-frame, therefore, we need a column that contains 150 data points. Let's create a vector `random_numbers` that holds 150 random numbers -

```{r}
set.seed(seed = 123)
random_numbers <- round(rnorm(150),2)
head(random_numbers)
length(random_numbers)
```

Now, we will add this vector to the data-frame as a column -
```{r}
# df_csv$random_number_column <- random_numbers
new_df <- cbind(df_csv, random_numbers)
head(new_df)

```

Similarly, let's create another vector of 6 matched data-type elements to match the six columns in the `new_df`, and add to the end of the data-frame -
```{r}
new_row <- c(mean(new_df$Sepal.Length), 
             mean(new_df$Sepal.Width),
             mean(new_df$Petal.Length),
             mean(new_df$Petal.Width),
             "random_sp",
             mean(new_df$random_numbers)
             )

new_df2 <- rbind(new_df, new_row)

tail(new_df2)

```

Now, we will touch on another important topic today - dealing with missing data. Missing data in *R* are represented by the symbol **NA** (not available) or impossible values (i.e. dividing by zero) are represented by the symbol **NaN** (not a number). Here we will deal with only **NA** which is the most common in day-to-day *R*. If we have a vector `x` which contains one or more **NA** values, we use `is.na()` function to inquire about **NA** values-
```{r}
x <- c(1,2,NA,3,4,NA,5,6)
is.na(x)
any(is.na(x))
```

The problem with the missing value(s) is that it hampers the simple operations, like - calculating the mean or sum etc -

```{r}
sum(x)
mean(x)
```
However, if you include a magic argument `na.rm=TRUE`, then the problem can be overcome and those operations can be performed on rest of the valid values of the vector `x` -
```{r}
sum(x, na.rm = TRUE)
mean(x, na.rm = TRUE)
```

However, a cleaver way to handle the missing data points is to *impute* them, and a clever way of doing is to take the mean of the rest of the data points -
```{r}
x[is.na(x)] <- mean(x, na.rm = T)
```

Obviously, there are more complex ways to impute the missing data, we will not cover those in this lecture

#### Short exercise on manipulating data


Please do the following steps -

1. Load the file named "iris_data_NA.csv".
2. Find out how many data points are missing.
3. Impute the missing data points with the median value of the respective columns. Round the values of the imputed data points to 2 decimal points using the funtion round()
4. What are the sums of each column (with numeric values)?


