# Control flow in *R* 


In this chapter, we will cover basically the so-called *Control flow in R*  with the following topics -

* Conditional execution
  + `If-else` statements
* Loops
  + `while` loop
  + `for` loop
* User defined funstions in *R*
  + Functions without `return`
  + Functions with `return`

## Conditional execution


### `If-else` statement

The very basic *conditional execution* is the `if` and `if-else` statements. Let's think you have a condition of do something (i.e. choose from a set of data-points or perform a mathematical operation) using *R*, you use , like -
```{r}
x <- 10
if(x == 10){
  print("x is equal to 10")
}
```

If you want to add more twist to the condition, you bring in `else` statement - `if(condition){do something} else{do the alternative}`. Like here -

```{r}
x <- 11
if(x == 10){
  print("x is equal to 10")
} else{
  print("x is not equal to 10")
}

```

You are not quiet content with the arguments here, you want to add more twists to the conditions, you add intermediate `else if` condition -
```{r}
x <- 11
if(x > 20){
  print("x is greater than 20")
} else if(x <= 20 & x >= 10){
  print("x is between 10 and 20")
} else{
  print("x is not less than  10")
}
```

### Short exercise

You have a job to grade 20 math scripts (the score ranges from 0 to 10) from a class of yr 7 students according to the following scale -

```
grade A: 8-10
grade B: 6-7
grade C: 4-5
grade D: 0-3

```

Now write a set of conditional `if else` statements so that if there are 100s of students, one can do the grading smoothly using your code. Now check if a student scored 5, what his/her grade would be?


## Loops

Loops are efficient ways to perform some mathematical operation iteratively, like - keep walking until you reach your destination (using *while loop*) or read each element of a vector, add one to it and do the same to the next (using *for loop*). 


### while loop

The basic structure of while loop is `while(condition){execute some action}`, like -


```{r}
# Think logically before executing the following code snippet -
# a <- 10
# while(a < 20){
#  print("a is less than 20")
# }
```

There are two ways to tackle this - 
1. you can add aother condition -
```{r}
a <- 10
while(a < 20){
  print("a is less than 20")
  a <- a + 1
  print(paste0("Because value of a is now ", a))
}

```

2. Using `break` to break out of the loop

```{r}
a <- 10
while(a < 20){
  
  a <- a + 1
  if(a < 20){
    print("a is less than 20")
    print(paste0("Because value of a is now ", a))
  }else{
    break
  }
  
}
```

Look carefully, though `break` is applied in the `if-else` statement, it is active to the loop that it is a part of.

### Short exercise

### for loop

Say, you have a vector `b <- c("a","b","c","d","e")` and you want to read the elements iteratively and then print them, what you can do is - 
```{r}
b <- c("a","b","c","d","e")

# looping through the contents of a vector directly
for(i in b){
  print(i)
}

# looping through the contents of a vector using indexing
for(i in 1:length(b)){
  print(b[i])
}

```

Looping through sequential numbers (in some programming language you have to define the initial value and the increment in this line) -
```{r}
for(i in 1:5){
  print(i)
}
```

For a bit better understanding -
```{r}
for(i in 1:5){
  print(paste0("The current value of i is ", i))
}
```

Look carefully at this piece of code to understand the nested loop and the change of the values of `i` and `j`

```{r}
for(i in 1:5){
  for(j in 6:10){
    print(paste("Now i is", i, "and j is", j))
  }
}
```

Or, the following code snippet -
```{r}

my_matrix <- matrix(1:18, nrow = 3, byrow = T)

for(i in my_matrix){
  print(i)
}

```

But you do not want to print it by column, rather you want to do it by row -

```{r}
for(row_index in 1:nrow(my_matrix)){
  for(col_index in 1:ncol(my_matrix)){
    print(my_matrix[row_index,col_index])
  }
}


```

### Short exercise

## User defined functions

By now, we have already seen some functions, like - `print()` and `mean()` that are built-in to your *R* environment. Sometimes they are not enough, you may need to perform some extra mathematical operations. Here comes the cool feature of *R* programming language, you can define your own functions. The basic structure of a fucntion is -

```
function_name <- function(input1, ..., inputN=default_value){
  code to execute
  return() # if any
}
```
### Function without return()

Let's first look at a very simple function. This does not take any input as prerequisit and will not return any value -

```{r}
printOnly <- function(){
  print("hi")
}
printOnly() # remember to put a parentheses afterwards to call the function.
```

Now, define the same function with an input as argument (with a default value as well) -

```{r}
printOnly <- function(name="David"){
  print(paste("hi", name))
}
printOnly()

# check out what happens if there is no default value assigned to the input argument and you call the function without any value.
```

### Function with a return()

So far, we have been, basically, printing something with the function. We actually want is to return the result. Now we will look into a function that returns a result and you can use it for later usage if you save it in a variable. Say, we want to add two numbers by a user defined function called `add` -

```{r}
add <- function(num1, num2){
  result <- num1 + num2
  return(result)
}

add(2,33)
```

Now, we can assign the output of the function to a variable (say, `result``) -

```{r}
add <- function(num1, num2){
  result <- num1 + num2
  return(result)
}

result <- add(2,33)
```

Check that nothing is printed on the console as it returned some value and the value is now saved in the variable `result`. If you type in `result` in the console, now you can see the value.

### Scoping

There is another interesting feature you should keep in mind while scripting and defining your own function(s) - The scoping. The value of a variable can change inside or outside of your function depending of it's position of definition and under what scope (either *global* or *local*) you call it -

```{r}

name <- "My name is David" # global scope
loc <- "I live in Edinburgh" # global scope

stay <- function(loc){
  print(name)
  loc <- "I live in London" # redefining locally and work within the scope of the function
  print(loc)
}
stay(loc)

# But, if you print "loc" now, see what happens......
print(loc)

```

## Exercise

